\chapter{Vocabulary}\label{chap:vocab}

P\=ali has plenty of words. In this digital age, finding a definition of P\=ali words is easy as other languages. Workable apps on your mobile phone or decent websites for this task are not hard to find. So, it is not sensible to list many of P\=ali terms here. However, learning to converse need some starting words applicable to our everyday life. There is no such a dictionary for this purpose. So, we have to collect some essential words to make our conversation effective. Grouping words into categories is also helpful. I take a number of words from A.\,P.\, Buddhadatta's \emph{Aids to Pali Conversation and Translation}.\footnote{\citealp{buddhadatta:aids}} This book already has a good number of words divided into groups. I do not transport all of them here. I ignore many words that we hardly use in common conversations. I also add words that I think they are useful in certain situations. As a result, our word bank is quite big. Moreover, I also give references to Abhidh\=anappad\=ipik\=a (only here abbreviated in tables as \emph{Abh.}), the traditional thesaurus, for the terms I can find some relations to the entries. This will help students to find alternatives and to dig deeper in traditional way.

\section{Nouns}\label{vocab:noun}
Nouns taken from the \emph{Aids} of Ven.\,Buddhadatta and newly added are reorganized. There are some adjectives that make sense to be annexed to noun tables. Some groups are overlapping. For example, if you can not find edible things in \emph{Food} category, try finding them in other possible group. We have around 850 nouns. The list of noun groups is shown in the table below.

\bigskip
\begin{longtable}[c]{%
	>{\raggedright\arraybackslash}p{0.52\linewidth}%
	>{\raggedleft\arraybackslash}p{0.25\linewidth}}
\caption*{Groups of nouns}\\
\toprule
\bfseries\upshape Group & \bfseries\upshape Page \\ \midrule
\endfirsthead
\multicolumn{2}{c}{Groups of nouns (contd\ldots)}\\
\toprule
\bfseries\upshape Group & \bfseries\upshape Page \\ \midrule
\endhead
\bottomrule
\ltblcontinuedbreak{2}
\endfoot
\bottomrule
\endlastfoot
%
Human beings & \pageref{vocabgrp1} \\
Parts of body & \pageref{vocabgrp2} \\
Occupations \& class of humans & \pageref{vocabgrp3} \\
Food & \pageref{vocabgrp4} \\
Fruits \& vegetables & \pageref{vocabgrp5} \\
Garments \& ornaments & \pageref{vocabgrp6} \\
Place-related terms & \pageref{vocabgrp7} \\
Vehicles & \pageref{vocabgrp8} \\
Parts of \& things in a house & \pageref{vocabgrp9} \\
Plant-related terms & \pageref{vocabgrp10} \\
Animal-related terms & \pageref{vocabgrp11} \\
Other nature-related terms & \pageref{vocabgrp12} \\
Time-related terms & \pageref{vocabgrp13} \\
Other common nouns & \pageref{vocabgrp14} \\
\end{longtable}

\begin{vocabNtable}{Human beings}\label{vocabgrp1}%
boy (child) & d\=araka, kum\=ara & m. & 253 \\
blood relation & s\=alohita & m. & 243 \\
brother & \texthl{bh\=atu}\footnote{Terms with color highlighted have irregular declension, see Appendix \ref{decl:irrn}, page \pageref{decl:pitu}.}, anuja & m. & 250 \\
brother-in-law & s\=ala & m. & 244 \\
\mbox{brother's daughter (niece)} & \texthl{bh\=atudh\=itu} & f. & \\
\mbox{brother's son (nephew)} & bh\=atuputta & m. & \\
daughter & \texthl{dh\=itu} & f. & 241 \\
daughter-in-law & su\d nis\=a & f. & 248 \\
elder brother & je\d t\d thabh\=atu & m. & 254 \\
elder sister & je\d t\d thabhagin\=i & f. & \\
family, clan & kula & nt. & 332 \\
father & \texthl{pitu} & m. & 243 \\
father-in-law & sasura & m. & 246 \\
father's sister & pitucch\=a & f. & 248 \\
friend & mitta, sakha, sah\=aya & m. & 346 \\
friend & sakh\=i & f. & 238 \\
girl (child) & d\=arik\=a, kum\=ar\=i & f. & 231 \\
girl, woman & ka\~n\~n\=a & f. & 231 \\
grandfather & ayyaka, pit\=amaha & m. & 247 \\
grandmother & ayyik\=a, m\=at\=amah\=i & f. & 245 \\
grandson & \texthl{nattu} & m. & 247 \\
great grandfather & payyaka & m. & 248 \\
great grandmother & payyik\=a & f. & \\
great grandson & \texthl{panattu} & m. & \\
husband & pati, \texthl{bhattu}, s\=am\=i & m. & 240 \\
husband's brother & devara & m. & 247 \\
infant & thanapa & m. & 252 \\
man, male & purisa & m. & 227 \\
mother & \texthl{m\=atu}, amm\=a & f. & 244 \\
mother-in-law & sassu & f. & 246 \\
mother's sister & m\=atucch\=a & f. & 248 \\
name & n\=ama & nt. & 114 \\
parents & m\=at\=apitu & m. & \\
person & puggala & m. & 93 \\
person, the people & jana & m. & 93 \\
old man & mahallaka & m. & 254 \\
old woman & mahallik\=a & f. & \\
relative, kinsman & \~n\=ati & m. & 243 \\
sister & bhagin\=i, anuj\=a & f. & 248 \\
sister-in-law & nanand\=a & f. & 245 \\
\mbox{sister's daughter (niece)} & bh\=agineyy\=a & m. & 246\\
\mbox{sister's son (nephew)} & bh\=agineyya & m. & \\
son, child & putta & m. & 240 \\
son-in-law & \texthl{j\=am\=atu} & m. & 247 \\
uncle (maternal) & m\=atula & m. & 245 \\
\mbox{uncle's wife (maternal)} & m\=atul\=an\=i & f. & 245 \\
uncle (paternal) & \texthl{c\=ulapitu} & m. & \\
\mbox{uncle's wife (paternal)} & \texthl{c\=ulam\=atu} & f. & \\
wife & bhariy\=a & f. & 237 \\
woman & itth\=i, n\=ar\=i, vanit\=a & f. & 230--1 \\
young man & taru\d na, yuv\=a\footnote{\pali{Yuv\=a} is nom.\ sg.\ of \pali{yuva}, see page \pageref{decl:yuva}. Sometimes \pali{yuva} is changed to \pali{yuv\=ana, yuna,} and \pali{y\=una} (Kacc\,649, R\=upa\,665, Sadd\,1287). These take regular paradigm of declension. In Kacc\,648, R\=upa\,664, \pali{taru\d na} can be changed to \pali{susu}, but in Sadd\,1286 both are seen as different words.} & m. & 252 \\
young woman & taru\d n\=i, yuvat\=i & f. & \\
younger brother & \texthl{ka\d ni\d t\d thabh\=atu}, anuja & m. & 254 \\
younger sister & ka\d ni\d t\d thabhagin\=i & f. & \\
\hline
\bfseries Related adjectives & & & \\
\hline
bald & khall\=a\d ta & & 321 \\
blind & andha & & 321 \\
blind (one-eyed) & k\=a\d na & & 321 \\
crippled & pa\.ngu & & 320 \\
deaf & badhira & & 322 \\
dumb, mute & m\=uga & & 320 \\
dwarfish & v\=amana & & 319 \\
humpbacked & khujja & & 319 \\
lame & kha\~nja & & 320 \\
squint-eyed & valira & & 320 \\
mad, insane & ummatta & & 322 \\
\end{vocabNtable}

\begin{vocabNtable}{Parts of body}\label{vocabgrp2}%
abdomen & vatthi & nt. & 276 \\
ankle & gopphaka & m. & 277 \\
anus & guda & nt. & 274 \\
arm & b\=ahu & m. & 265 \\
armpit & kaccha & m. & 264 \\
back & pi\d t\d thi & f. & 260 \\
belly & kucchi & m. f. & 271 \\
blood & lohita & nt. & 280--1 \\
body & k\=aya & m. & 151 \\
bone & a\d t\d thi & nt. & 278 \\
bowel & anta & nt. & 271 \\
brain & matthalu\.nga & nt. & \\
breast, chest & ura & m. & 270 \\
breast of a woman & thana & m. & 270 \\
cheek & kapola & m. & 262 \\
chin & cubuka & nt. & 262 \\
chin & hanu & f. & 262 \\
ear & ka\d n\d na & m. & 150 \\
elbows & kappara & m. & 265 \\
excrement & g\=utha, kar\=isa, vacca & nt. & 274--5\\
eye & akkhi, cakkhu & nt. & 149 \\
eyebrow & bhamu, bhamuk\=a & f. & 259 \\
eyelash & pakhuma & nt. & 259 \\
face & \mbox{\=anana, mukha, vadana} & nt. & 260\footnote{Face and mouth use the same set of words.} \\
fang & d\=a\d th\=a & f. & 261 \\
fat & meda & m. & 282 \\
finger & a\.nguli & f. & 266 \\
flesh & ma\d msa & nt. & 280 \\
foot, leg & p\=ada & m. & 277 \\
hand & hattha, p\=a\d ni & m. & 265 \\
hair (head) & kesa & m. & 257 \\
hair (body) & loma & nt. & 259 \\
head & s\=isa & nt. & 256 \\
heart & hadaya & nt. & 270\footnote{In Abhidh\=a heart and breast share the same meaning.} \\
heel & pa\d nhi & f. &  277 \\
hip & ka\d ti & f. & 272 \\
hip & jaghana & nt. & 272 \\
jaw & hanuk\=a & f. & 262 \\
kidney & vakka & nt. & \\
knee & j\=anu & m. & 276 \\
lip & o\d t\d tha & m. & 262 \\
liver & yakana & nt. & \\
lungs & papph\=asa & nt. & \\
mouth & mukha & nt. & 260 \\
nail & nakha & m. & 268 \\
neck & g\=iv\=a & f. & 263 \\
nerve & sir\=a & f. & 279 \\
nose & gh\=a\d na & nt. & 150 \\
nose & n\=as\=a, n\=asik\=a & f. & 150 \\
organ & a\.nga & nt. & 278 \\
organ & avayava & m. & 278 \\
pus & p\=uya & m. & 325 \\
rib & ph\=asuk\=a & f. & 278 \\
saliva & khela & m. & 281 \\
semen & sukka & nt. & 274 \\
shoulder & a\d msak\=u\d ta & m. & 264 \\
sinew, tendon & nh\=ar\=u\footnote{Also \pali{nah\=ar\=u} can be found, as well as the short ending (\pali{-u}) version of the both terms.} & m. & 279 \\
skin & taca & m. & \\
spleen & pihaka & nt. & \\
stomach & udara & nt. & 271 \\
sweat & seda & m. & \\
tear & assu & nt. & 260 \\
thigh & \=uru & m. & 276 \\
thumb & a\.ngu\d t\d tha & m. & 266 \\
toe & p\=ada\.nguli & f. & \\
tooth & danta & m. & 261 \\
urine & mutta & nt. & 275 \\
\end{vocabNtable}

\begin{vocabNtable}{Occupations \& class of humans}\label{vocabgrp3}%
artisan, craftman & sippaka, sipp\=i & m. & 504 \\
artist & cittak\=ara & m. & 508 \\
author & ganthak\=ara & m. & \\
baker & p\=upak\=ara & m. & \\
barber & nah\=apita & m. & 505 \\
beggar & y\=acaka & m. & 740 \\
blacksmith & kamm\=ara & m. & 509 \\
Buddhist monk & bhikkhu & m. & 433 \\
captain, commander & niy\=amaka & m. & 667 \\
carpenter & va\d d\d dhak\=i & m. & 506 \\
clerk & lekhaka & m. & 348 \\
cook & s\=uda & m. & 464 \\
cripple & p\=i\d thasapp\=i & m. & 319 \\
customer & kayika & m. & 470 \\
dancer & na\d ta & m. & 101 \\
doctor & vejja & m. & 329 \\
driver, charioteer & s\=arathi, rath\=acariya & m. & 376 \\
enemy, foe & ari, sapatta & m. & 344--5 \\
engineer & yantasipp\=i & m. & \\
farmer & kassaka & m. & 447 \\
fisherman & keva\d t\d ta & m. & 670 \\
fool & b\=ala, mu\d lha & m. & 721 \\
foreigner & vides\=i, vij\=atika & m. & \\
gamester, gambler & j\=utak\=ara & m. & 531 \\
gardener & uyy\=anap\=ala & m. & \\
goldsmith & suva\~n\~nak\=ara & m. & 506 \\
guest & atithi, \=agantuka & m. & 424 \\
hair-dresser & nah\=apita & m. & 505 \\
hermit & t\=apasa, isi & m. & 433 \\
hunter & vy\=adha, ludda & m. & 518 \\
keeper, guard & p\=ala & m. & \\
lawyer & n\=itived\=i & m. & \\
leader & n\=ayaka, s\=am\=i, pati & m. & 725 \\
learner & sikkhata, sekkha & m. & \\
mathematician & ga\d naka & m. & 347 \\
mechanic & yantika & m. & \\
merchant & v\=a\d nija & m. & 469 \\
millionaire & se\d t\d th\=i, a\d d\d dha, dhan\=i & m. & 725 \\
miser & kapa\d na & m. & 739 \\
monk & muni, sama\d na & m. & 433 \\
musician & v\=adaka & m. & \\
mystic & yog\=i & m. & \\
nun & sama\d n\=i & f. & \\
nurse & dh\=at\=i & f. & \\
painter & va\~n\~n\=alepaka & m. & \\
pilot & ka\d n\d nadh\=ara & m. & \\
police officer & bha\d ta & m. & 376 \\
poor person & da\d lidda, dalidda & m. & 739 \\
preacher & desaka & m. & \\
publisher & pak\=asaka & m. & \\
sailor & n\=avika & m. & 666 \\
scribe & lekhaka & m. & 348 \\
seller & vikkayika & m. & 470 \\
servant, attendant & sevaka & m. & 342 \\
shopkeeper & \=apa\d nika & m. & 469 \\
singer & g\=ayaka & m. & \\
soldier & yodha & m. & 376 \\
student & sissa, antev\=as\=i & m. & 408 \\
tailor & tu\d n\d nav\=aya & m. & \\
teacher & \=acariya & m. & 410 \\
thief & cora & m. & 522 \\
warrior & khattiya & m. & 335 \\
washerman & rajaka, dhovaka & m. & \\
widow & vidhav\=a & f. & 235 \\
widower & matabhariya & m. & \\
wise man & pa\d n\d dita & m. & 227--9 \\
worker & kammak\=ara, kammakara & m. & 514 \\
\end{vocabNtable}

\begin{vocabNtable}{Food}\label{vocabgrp4}%
barley & yava & m. & 450 \\
bean & m\=asa & m. & \\
beef & goma\d msa & nt. & \\
beverage & p\=ana, p\=anaka & nt. & \\
bread & godhumap\=upa & m. & \\
butter & non\=ita & nt. & 500 \\
butter milk & takka & nt. & 500 \\
cake & p\=uva, p\=upa & m. & 463 \\
candy & kha\d n\d da & m. & 462 \\
chillies & marica & nt. & 459 \\
coffee & k\=aph\=i & f. & \\
corn & varaka & m. & 450 \\
curd & dadhi & nt. & 501 \\
curry & s\=upa & m. & 464 \\
curry & bya\~njana (vya\~njana) & nt. & 464 \\
fat & vas\=a & f. & 282 \\
fish & maccha, m\=ina & m. & 671 \\
dried fish & sukkhamaccha & m. & \\
salted fish & lo\d namaccha & m. & \\
flour & pi\d t\d tha & nt. & 1075 \\
fried flour & sattu & m. & 1122 \\
food & \=ah\=ara & m. & 465 \\
food & bhojana & nt. & 465 \\
hard food & kh\=adan\=iya & nt. & \\
soft food & bhojan\=iya & nt. & \\
garlic & lasu\d na & nt. & 595 \\
ghee & ghata & nt. & 501 \\
grain & dha\~n\~na & nt. & 552 \\
green peas & mugga & m. & 450 \\
honey & madhu & nt. & 494 \\
kernel & mi\~nj\=a & f. & 593 \\
liquor (distilled) & sur\=a & f. & 533 \\
liquor (fermented) & meraya & nt. & 533 \\
meat & ma\d msa, \=amisa & nt. & 280 \\
dried meat & vall\=ura & nt. & 280 \\
milk & kh\=ira & nt. & 500 \\
molasses, treacle & ph\=a\d nita & nt. & 462 \\
mustard & siddhattha & m. & 451 \\
mutton & e\d lakama\d msa & nt. & \\
oil & tela & nt. & \\
onion & pala\d n\d du & m. & 595 \\
paddy & v\=ihi & m. & 452 \\
pepper & kolaka & nt. & 304 \\
pork & s\=ukarama\d msa & nt.& \\
rice grain & ta\d n\d dula & nt. & \\
boiled rice & odana & m. & 465 \\
boiled rice & bhatta & nt. & 465 \\
rice gruel & y\=agu & f. & 465 \\
salad & s\=alava & m. & \\
salt & lava\d na, lo\d na & nt. & 460 \\
sesame & tila & nt. & \\
sugar & sakkhar\=a & f. & 462 \\
sweet, candy & khajjaka & nt. & \\
syrup & p\=ana, p\=anaka & nt. & \\
tamarind & ci\~nc\=a & f. & 562 \\
tea & c\=ah\=a & f. & \\
turmeric & ha\d lidd\=a, ha\d lidd\=i & f. & 586 \\
vegetable & \mbox{harita, haritaka, pa\d n\d na} & nt. & \\
venison & migama\d msa & nt. & \\
vinegar & bila\.nga & m. & \\
wine & muddik\=asava & m. & \\
yam & kanda, \=aluva & m. & \\
\hline
\bfseries Related adjectives & & & \\
\hline
boiled & sedita & & \\
cooked & pakka, pacita & & \\
fried & bhajjita & & \\
roasted & a\.ng\=arapakka & & \\
bitter & titta & & \\
pungent & ka\d tuka & & \\
salty & lava\d na & & \\
sour & ambila & & \\
sweet & madhura & & \\
\end{vocabNtable}

\begin{vocabNtable}{Fruits \& vegetables}\label{vocabgrp5}%
banana & kadaliphala & nt. & 589 \\
bitter gourd & k\=aravella & m. & 596 \\
breadfruit & labuja & m. & 570 \\
brinjal, eggplant & v\=ati\d.ngana, bha\d n\d d\=ak\=i & m. & 588 \\
cabbage & go\d lapatta & nt. & \\
coconut & n\=a\d likera & nt. & 604 \\
cucumber, pumpkin & kakk\=ar\=i & f. & 597 \\
dates & khajj\=ur\=i & f. & 603--4 \\
fig & udumbara & nt. & 551 \\
fruit & phala & nt. & \\
gourd & l\=abu & nt. & 596 \\
snake gourd & pa\d tola & nt. & 595 \\
grape & muddik\=a & f. & 597 \\
horseradish & siggu & nt. & 554 \\
jujube & badar\=i & f. & 558 \\
jujube & badara & m. nt. & 559 \\
jackfruit & panasa & nt. & 569 \\
lotus root & mu\d l\=ala & nt. & 687 \\
mango & amba & nt. & 557 \\
mangosteen & madhutimbaru & nt. & \\
orange & jamb\=ira, n\=ara\.nga & m. & 560 \\
papaya & v\=atakumbhaphala & nt. & \\
pineapple & madhuketak\=i & m. & \\
pineapple & bahunettaphala & nt. & \\
pumpkin, gourd & kumbha\d n\d da & m. & 597 \\
radish & m\=ulaka & m. & 598 \\
rhubarb & tambaka & m. & 598 \\
roseapple & jamb\=u & f. & 547 \\
sugarcane & ucchu & nt. & 599 \\
watermelon & vallibha & m. & 597 \\
wood apple & kapittha & nt. & 551 \\
\hline
\bfseries Related adjectives & & & \\
\hline
fresh & abhinava & & \\
ripe & pakka & & \\
unripe & \=amaka & & \\
rotten & p\=utika & & \\
\end{vocabNtable}

\begin{vocabNtable}{Garments \& ornaments}\label{vocabgrp6}%
bangle & valaya & nt. & 285 \\
bangle for arm & key\=ura & nt. & 287 \\
bangle for feet & n\=upura & m. & 288--9 \\
belt & k\=ayabandhana & nt. & \\
blanket & kambala & nt. & 298 \\
bracelet & ka\d taka, valaya & nt. & 285 \\
calico & kapp\=asika & nt. & 297 \\
chaplet & sekhara & m. & 308 \\
cloth & vattha, dussa & nt. & 290 \\
cloth for bathing & udakas\=a\d tik\=a & f. & \\
collar & g\=iveyya & nt. & 285 \\
crest gem, diadem & c\=u\d l\=ama\d ni & m. & 283 \\
crown & kir\=i\d ta & nt. & 283 \\
earring & ku\d n\d dala & nt. & 284 \\
garland & m\=al\=a & f. & 307 \\
garment & s\=a\d taka & nt. & 290 \\
inner garment & antarav\=asaka & m. & 292 \\
lower garment & niv\=asana & nt. & 292 \\
upper garment & uttar\=asa\.nga & m. & 292--3 \\
upper garment & uttar\=iya & nt. & 293 \\
girdle & ka\d tibandhana & nt. & \\
girdle of a woman & mekhal\=a, rasan\=a & f. & 287 \\
handkerchief & hatthapu\~nchana & nt. & \\
hat & n\=a\d lipa\d t\d ta & m. & 294 \\
hem & das\=a & f. & 294 \\
jacket & ka\~ncuka & m. & 294 \\
necklace & h\=ara & m. & 285 \\
linen & khoma & nt. & 297 \\
ointment & vilepana & nt. & 306 \\
overcoat & d\=ighaka\~ncuka & m. & \\
perfume & sugandha & m. & 146 \\
perfuming & v\=asana & nt. & 307 \\
pin & sal\=ak\=a & f. & \\
rag & kappa\d ta & m. & 293 \\
rag & nantaka & nt. & 293 \\
ring & a\.ngul\=iyaka & nt. & 286 \\
ring & a\.ngulimudd\=a & f. & 287 \\
robe & c\=ivara & nt. & 296 \\
signet ring & muddik\=a & f. & 287 \\
sandals & up\=ahana & nt. & 525 \\
scent & gandhas\=ara & m. & \\
shoe & p\=aduk\=a, p\=adu & f. & 525 \\
silk & koseyya & nt. & 291 \\
smoking pipe & dh\=umanetta & nt. & \\
soap & nah\=an\=iya & nt. & \\
string of pearls & mutt\=avali & f. & 285 \\
towel & mukhapu\~nchana & nt. & \\
turban & s\=isave\d thana & nt. & \\
umbrella & chatta & nt. & 357 \\
veil & mukh\=avara\d na & nt. & \\
walking stick & kattaraya\d t\d tthi & f. & \\
watch & hor\=alocana & nt. & \\
woolen & kambala & nt. & 298 \\
\end{vocabNtable}

\begin{vocabNtable}{Place-related terms}\label{vocabgrp7}%
area, place & desa, visaya & m. & 186 \\
bank & dhan\=ag\=ara & nt. & \\
barber's shop & nah\=apitas\=al\=a & f. & \\
barn & ko\d t\d tha & nt. & 458 \\
barrack & yodhaniv\=asa & m. & \\
bridge & setu & m. & 189 \\
bus terminal & mah\=arathos\=ana & nt. & \\
castle, mansion & p\=as\=ada & m. & 208 \\
cave & guh\=a & f. & 609 \\
cemetery & sus\=ana & nt. & 405 \\
city, town & nagara, pura & nt. & 198 \\
college & vijj\=alaya & m. & \\
continent & mah\=ad\=ipa & m. & \\
country & ra\d t\d tha & nt. & 189 \\
countryside & paccanta & m. & 186 \\
crossroad & maggasandhi & f. & \\
factory & kammantas\=al\=a & f. & \\
field, arable land & ked\=ara, khetta & nt. & 447 \\
flower garden & pupph\=ar\=ama & m. & \\
foreign country & vedesa & m. & \\
garage & rathas\=al\=a & f. & \\
granary & kus\=ula & m. & 458 \\
granary & dha\~n\~n\=ag\=ara & nt. & \\
hall, shed & s\=al\=a & f. & 207 \\
harbor, port & tittha & nt. & \\
hospital & \=arogyas\=al\=a & f. & \\
hotel & bhojan\=ag\=ara & nt. & \\
house & \=av\=asa, niv\=asa, \=alaya & m. & 205--7 \\
house & geha, ghara, ag\=ara & nt. & 205--7 \\
hut & ku\d ti, pa\d n\d nas\=al\=a & f. & \\
jail & bandhan\=ag\=ara & nt. & 407 \\
law court & vinicchayas\=al\=a & f. & \\
library & potthak\=alaya & m. & \\
lunatic asylum & ummatt\=alaya & m. & \\
market & \=apa\d na & m. & 213 \\
market town & nigama & m. & 225 \\
metropolis & r\=ajadh\=ani & f. & 198 \\
mine & \=akara & m. & \\
monastery & \=ar\=ama & m. & \\
orchard & phal\=ar\=ama & m. & \\
park & \=ar\=ama & m. & 537 \\
park & uyy\=ana & nt. & 538 \\
pharmacy & bhesajj\=ag\=ara & nt. & \\
photographic studio & ch\=ay\=ar\=upas\=al\=a & f. & \\
place & \d th\=ana & nt. & 846 \\
place, district & padesa & m. & \\
police station & bha\d t\=ag\=ara & nt. & \\
post office & sandes\=ag\=ara & nt. & \\
printing office & mudda\.nka\d n\=alaya & m. & \\
railway station & \mbox{dh\=umarathanivattana} & nt. & \\
restaurant & lahubhojan\=ag\=ara & nt. & \\
school & p\=a\d thas\=al\=a, sippas\=al\=a & f. & 212 \\
site for building & gharavatthu & nt. & \\
street & racch\=a, v\=ithi & f. & 202 \\
tavern & \=ap\=ana, p\=anama\d n\d dala & nt. & 534 \\
temple & vih\=ara, \=asama & m. & \\
theater & naccas\=al\=a & f. & \\
toilet & vaccaku\d ti, pass\=avaku\d ti & f. & 212 \\
university & mah\=avijj\=alaya, nikhilavijj\=alaya & m. & \\
upper floor & p\=as\=adatala & nt. & \\
village & g\=ama & m. & 225 \\
warehouse & bha\d n\d d\=ag\=ara & nt. & \\
\hline
\bfseries Related adjectives & & & \\
\hline
northern, higher & uttara & & \\
southern, right & dakkhi\d na & & \\
eastern & puratthima & & \\
western & pacchima & & \\
lower & adhara & & \\
left & v\=ama & & \\
central & majjhima & & \\
bordering & paccanta & & 186 \\
\end{vocabNtable}

\begin{vocabNtable}{Vehicles}\label{vocabgrp8}%
aeroplane & \mbox{gaganay\=ana, vyomay\=ana} & nt. & \\
canoe & khuddakado\d ni & f. & \\
car, carriage, vehicle & ratha & m. & 372 \\
boat & do\d ni & f. & 668 \\
bus & mah\=aratha & m. & \\
ferry & tittha & nt. & \\
oar & aritta & nt. & 667 \\
raft & u\d lumpa & m. & 665 \\
rudder & ka\d n\d na & nt. & \\
sail & lak\=ara & m. & \\
ship & n\=av\=a & f. & 666 \\
submarine & antodakan\=av\=a & f. & \\
train & dh\=umaratha & m. & \\
tram & vijjuratha & m. & \\
vehicle & y\=ana, v\=ahana & nt. & 375 \\
\end{vocabNtable}

\begin{vocabNtable}{Parts of \& things in a house}\label{vocabgrp9}%
bag, purse & pasibbaka & m. & \\
balcony, veranda & \=alinda & m. & 218 \\
basket & pi\d taka & nt. & 524 \\
bench & nis\=idanaphalaka & m. & \\
board, plank & phalaka & m. & \\
book & potthaka & m. nt. & 1006 \\
bottle & n\=a\d lik\=a & f. & \\
glass bottle & k\=acatumba & m. & \\
bed & ma\~nca & m. & 308 \\
bedroom & sayanighara & nt. & \\
broom & sammajjan\=i & f. & 223 \\
brush & v\=ala\d n\d d\=upaka & m. & \\
carpet & kojava & m. & 312 \\
casket & samugga & m. & 317 \\
chair & p\=i\d tha & nt. & 311 \\
easy chair & \=asandi & f. & \\
couch & palla\.nka & m. & 308 \\
cup & casaka & m. & 534 \\
dining room & bhojan\=ag\=ara & nt. & \\
dining table & bhojanaphalaka & m. & \\
dish & th\=ali & f. & 456 \\
door & dv\=ara & nt. & 219 \\
fan & b\=ijan\=i (v\=ijan\=i) & f. & 316 \\
fire place & uddhana & nt. & \\
firewood & d\=aru, indhana & nt. & 547 \\
flag & ketu, dhaja & m. & 397 \\
furniture & d\=arubha\d n\d da & nt. & \\
glassware & k\=acabha\d n\d da & nt. & \\
handbasket & pacchi & f. & 524 \\
jar (big) & ara\~njara & m. & 456 \\
jug & ku\d n\d dik\=a & f. & 443 \\
key & ku\~ncik\=a & f. & 222 \\
key & t\=a\d la & m. & 222 \\
key hole & ku\~ncik\=avivara & nt. & 222 \\
knife, dagger & ch\=urik\=a, churik\=a & f. & 392 \\
knife, lance & sattha & nt. & 385 \\
lamp & pad\=ipa & m. & 316 \\
lampwick & d\=ipava\d t\d ti & f. & \\
lavatory & vaccaku\d ti & f. & 212 \\
leathern bag & bhast\=a & f. & \\
mat & kila\~nja & m. & 455 \\
mirror & \=ad\=asa & m. & 316 \\
mosquito net & makas\=avara\d na & nt. & \\
needle & s\=uci & m. & \\
net & j\=ala & nt. & 949 \\
newspaper & pavattipatta & nt. & \\
paper & pa\d n\d na & nt. & \\
pen & lekhan\=i & f. & \\
pencil & abbhakalekhan\=i & f. & \\
pillow & bimbohana & nt. & 311 \\
plate (to eat from) & ka\d msa & nt. & 905 \\
pot & c\=a\d ti, kumbh\=i & f. & 456 \\
roof & chadana & nt. & 218 \\
room & gabbha, ovaraka & m. & 214 \\
rope & rajju & f. & 448 \\
ruler & ujurekhaka & m. & \\
scales & tul\=a & f. & 823 \\
scissors & kattarik\=a & f. & \\
seat & \=asana & nt. & 311 \\
sewing machine & sibbanayanta & nt. & \\
spoon & ka\d tacchu & m. & 458 \\
stair & sop\=a\d na & nt. & 216 \\
table & phalakadh\=ara & m. & \\
telephone & d\=urabh\=asanayanta & nt. & \\
terrace & vedik\=a & f. & \\
thread & sutta & nt. & 523 \\
thread & tantu & m. & 523 \\
tray & ka\d ms\=adh\=ara & m. & \\
vessel & bh\=ajana & nt. & 457 \\
wall & bhitti & f. & 204 \\
waterpot & gha\d ta & m. & 457 \\
water tank & jal\=asaya & m. & \\
window & v\=atap\=ana & nt. & 216--7 \\
window shutter & kav\=a\d ta & nt. & \\
writing desk & lekhanaphalaka & m. & \\
\end{vocabNtable}

\begin{vocabNtable}{Plant-related terms}\label{vocabgrp10}%
banyan & nigrodha & m. & 551 \\
bamboo & ve\d lu, ve\d nu & m. & 599--600\\
branch & s\=akh\=a & f. & 542 \\
bud & makula & nt. & 544 \\
bush, thicket & gumba & m. & 861 \\
cidar & khadira & m. & 567 \\
coconut palm & n\=a\d likera & m. & 604 \\
date palm & khajj\=ur\=i & f. & 603--4 \\
ebony & k\=a\d las\=ara & m. & \\
fig tree & udumbara & m. & 551 \\
holy fig, bo & assattha & m. & 551 \\
flower & puppha, kusuma & nt. & 545 \\
fruit & phala & nt & \\
grapevine & muddik\=a & f. & \\
leaf & patta, pa\d n\d na & nt. & 543 \\
lotus & paduma & nt. & 684--5 \\
white lotus & pu\d n\d dar\=ika & nt. & 686 \\
mango tree & ambarukkha & m. & 557 \\
reed & na\d la & m. & 601 \\
root & m\=ula & nt. & 549 \\
sal tree & s\=ala & m. & 562 \\
shrub & gaccha & m. & 540 \\
shrub & la\d t\d thi, la\d t\d thik\=a & f. & \\
sprout & a\.nkura & m. & 543 \\
stalk & va\d n\d ta & nt. & 544 \\
talipot palm & t\=ala, t\=alarukkha & m. & 604 \\
talipot palm & t\=al\=i & f. & 604 \\
tree & rukkha, taru & m. & 539--40 \\
trunk & khanda & m. & \\
twig & pas\=akh\=a & f. & \\
vine & lat\=a & f. & 550 \\
water lily & uppala & nt. & 688 \\
white water lily & kumuda & nt. & 688 \\
\end{vocabNtable}

\begin{vocabNtable}{Animal-related terms}\label{vocabgrp11}%
animal & tiracch\=ana & m. & 648 \\
ant & pip\=ilik\=a, kapillik\=a & f. & \\
anthill & vammika & m. & \\
bat (large) & jatuk\=a & f. & 646 \\
bat (small) & vagguli & f. & \\
bear & accha & m. & 612 \\
beast & pasu & m. & 1111 \\
bee & madhukara, ali & m. & 635--6 \\
bird & pakkhina & m. & 624--5 \\
bitch & sunakh\=i & f. & \\
boa & ajagara & m. & 651 \\
buffalo & mahisa & m. & 616 \\
bug & ma\.nku\d na & m. & \\
calf & vaccha, potaka & m. & \\
camel & o\d t\d tha & m. & 502 \\
cat & bi\d l\=ala & m. & 615 \\
cattle & \texthl{go}\footnote{\pali{Go} has irregular declension, see Appendix \ref{decl:irrn}, page \pageref{decl:go}.} & m. & \\
centipede & satapad\=i & m. & 622 \\
chameleon & kaka\d n\d taka & m. & 623 \\
cobra & n\=aga & m. & \\
cock & kukku\d ta & m. & 640 \\
conch & sa\.nkha & m. & 676 \\
cow & g\=av\=i & f. & 498 \\
crab & ku\d l\=ira & m. & 675 \\
crane & baka & m. & 642 \\
cricket & c\=ir\=i & f. & 646 \\
crocodile & susum\=ara, kumbh\=ila & m. & 674 \\
crow & k\=aka & m. & 638 \\
crow & k\=ak\=i & f. & \\
deer & miga & m. & 620 \\
deer & mig\=i & f. & \\
dog & sunakha & m. & 518--9 \\
donkey & gadrabha & m. & \\
dove & kapota & m. & 636 \\
drake & k\=adamba & m. & 644 \\
eagle & garu\d la & m. & 633 \\
egg & a\d n\d da, b\=ija & nt. & 1092 \\
elephant & hatth\=i, ku\~njara, gaja & m. & 360 \\
elephant & hatthin\=i & f. & 362 \\
feather & pekhu\d na, patta & nt. & 627 \\
female animal & dhenu & f. & 498 \\
female bird & pakkhidhenu & f. & \\
flea & upp\=ataka & m. & \\
fly & makkhik\=a & f. & \\
fox & sig\=ara & m. & 615 \\
frog & ma\d n\d d\=uka & m. & 675 \\
gecko & sarab\=u & f. & 621 \\
goat & e\d laka, aja & m. & 502 \\
hare & sasa & m. & 617 \\
hawk & sena & m. & 637 \\
herd & y\=utha & m. & 632 \\
heron & ka\.nka & m. & 643 \\
hide & camma & nt. & 442 \\
hoof & khura & m. & 371 \\
horn & vis\=a\d na, si\.nga & nt. & 497 \\
hump & kaku & m. & 497 \\
hen & kukkut\=i & hen & \\
horse & assa, haya & m. & 368 \\
house lizard & gharago\d lik\=a & f. & 621 \\
hyena & taraccha & m. & 611 \\
iguana, lizard & godh\=a & f. & 622 \\
insect & k\=i\d ta & m. & 623 \\
leech & jal\=uk\=a & f. & 675 \\
leopard, panther & d\=ipi & m. & \\
lion & s\=iha & m. & \\
lioness & s\=ih\=i & f. & \\
louse & \=uk\=a & f. & \\
monkey & v\=anara & m. & 613--4 \\
mosquito & makasa & m. & 646 \\
nest & kul\=avaka & m. & 627 \\
owl & u\d l\=uka, ul\=uka & m. & 638 \\
ox, bull & go\d na, usabha & m. & 495 \\
oyster & sippik\=a & f. & \\
parrot & suka, sava, suva & m. & 640 \\
peacock & may\=ura, mora & m. & 634 \\
pig & s\=ukara & m. & 617 \\
pigeon & p\=ar\=avata & m. & 636 \\
porcupine & salla & m. & 616 \\
quadruped & catuppada & m. & \\
ram & me\d n\d da & m. & 501 \\
rat & \=akhu, m\=usika & m. & 618 \\
rhinoceros & khaggavis\=a\d na & m. & 613 \\
scorpion & vicchik\=a & f. & 621 \\
serpent, snake & sappa, ahi & m. & 653--4 \\
shark & makara & m. & \\
sparrow & c\=ataka & m. & 641 \\
spider & u\d n\d nan\=abhi & m. & 621 \\
squirrel & kalandaka & m. & 622 \\
swan & ha\d msa & m. & 646 \\
termite & upacik\=a & f. & \\
tiger & byaggha (vyaggha) & m. & 611 \\
tortoise & kumma & m. & 674 \\
turtle & kacchapa & m. & 674 \\
venom & visa & nt. & 655 \\
viper & gonasa & m. & 651 \\
vulture & gijjha & m. & 637 \\
wing & pakkha & m. & 627 \\
wolf & vaka & m. & 615 \\
worm & kimi, k\=i\d ta & m. & 623 \\
\end{vocabNtable}

\begin{vocabNtable}{Other nature-related terms}\label{vocabgrp12}%
brass & \=arak\=u\d ta & m. & 492 \\
bronze & ka\d msa & m. & 905 \\
canal & jalam\=atik\=a & f. & \\
cliff & pap\=ata, ta\d ta & m. & 608 \\
cloud & val\=ahaka & m. & 47--8 \\
comet & dh\=umaketu & m. & \\
constellation & nakkhatta & nt. & 57 \\
copper & tamba & m. & 964 \\
coral & pav\=ala & nt. & 491 \\
cyclone & verambhav\=ata & m. & \\
darkness & andhak\=ara & m. & 70 \\
desert & kant\=ara & m. & 192 \\
dew, snow & tuhina & nt. & 56 \\
diamond & vajira & nt. & 490 \\
dust & dh\=uli & f. & 395 \\
earth & pa\d thav\=i & f. & 181 \\
element & dh\=atu & f. & 817 \\
emerald & marakata & nt. & \\
fire & aggi & m. & 33--4 \\
flood & ogha & m. & 947 \\
forest & ara\~n\~na, vana & nt. & 536 \\
gem, jewel & ma\d ni & m. & 489 \\
gold & suva\d n\d na & nt. & 487--8 \\
gold bullion & hira\~n\~na & nt. & 486--8 \\
ground & bh\=umi & f. & 1098 \\
heat & u\d nha & nt. & \\
ice, snow & hima & nt. & 56 \\
iron & \texthl{aya}\footnote{\pali{Aya} has irregular declension as \pali{mana} group, see page \pageref{decl:mana}.} & m. & 493 \\
island & d\=ipa & m. & 664 \\
lagoon & lo\d n\=i & f. & \\
lake & sara, ta\d l\=aka & m. & 678 \\
land & thala & nt. & 664 \\
lead & tipu & nt. & \\
light & \=aloka & m. & 37 \\
lightning & vijju & f. & 48 \\
lotus pond & ambujjin\=i & f. & 689 \\
mercury & p\=arada & m. & 493 \\
metal & loha & m. nt. & 493 \\
meteor & ukk\=ap\=ata & m. & \\
mineral & kha\d nija & nt. & \\
moon & canda & m. & 51--2 \\
moonlight & candik\=a & f. & 54 \\
mountain & pabbata & m. & 605 \\
mountaintop & k\=u\d ta & m. & 608 \\
mud & pa\.nka & m. & 663 \\
mud & kalala & nt. & 663 \\
ocean, sea & samudda, s\=agara & m. & 659 \\
pearl & mutt\=a & f. & 492 \\
planet & gahat\=ar\=a & f. & \\
pond & pokkhara\d n\=i & f. & 677 \\
rain & megha\footnote{rainy cloud} & m. & 47 \\
rain & vassa & nt. & 48 \\
rainbow & indadhanu & nt. & 49 \\
river & nad\=i & f. & 681 \\
river's mouth & nad\=imukha & nt. & \\
river's mouth & sindhusa\.ngama & m. & 681 \\
riverbank & k\=ula, tira & nt. & 664 \\
further bank & p\=ara & nt. & 665 \\
near bank & ora & nt. & 665 \\
ruby & lohit\.nka & m. & 491 \\
sand & sikat\=a, v\=alik\=a & f. & 663 \\
sapphire & indan\=ila & m. & \\
shade, shadow & ch\=ay\=a & f. & 954 \\
shore & vel\=a & f. & 660 \\
silver & rajata & nt. & 489 \\
sky & \=ak\=asa & m. & 45--6 \\
sky & nabha & nt. & 45--6 \\
smoke & dh\=uma & m. & \\
soil & mattik\=a & f. & 182 \\
the solar system & cakkav\=a\d la & m. & \\
star & t\=ar\=a, t\=arak\=a & f. & 57 \\
stone, rock & sela, p\=as\=a\d na & m. & 605 \\
stone, rock & sil\=a & f. & 605 \\
stream & kunnad\=i & f. & \\
sun & suriya, \=adicca, ravi & m. & 62--3 \\
swamp & anupa & m. & \\
thunder & thanita, dhanita & nt. & 49 \\
thunderbolt & asani & f. & 871 \\
topaz & phussar\=aga & m. & \\
torrent & jaladh\=ar\=a & f. & \\
universe & sakalalokadh\=atu & f. & \\
valley & upaccak\=a & f. & 610 \\
water & ambu, udaka, jala & nt. & 661 \\
waterfall, cascade & nijjhara & m. & 608 \\
wave & \=umi & f. & 662 \\
wave & tara\.nga & m. & 662 \\
well & k\=upa & m. & 931 \\
whirlpool & \=ava\d t\d ta & m. & 660 \\
wind & v\=ata, v\=ayu & m. & 37 \\
world & loka & m. & 186 \\
\end{vocabNtable}

\begin{vocabNtable}{Time-related terms}\label{vocabgrp13}%
time & k\=ala & m. & 66 \\
hour & gha\d tik\=a & f. & 74 \\
moment & kha\d na & m. & 855 \\
occasion, turn & v\=ara & m. & 1042 \\
yesterday & h\=iyo, hiyyo & (ind.) & 1155 \\
today & ajja & (ind.) & 1155 \\
tomorrow & sve, suve & (ind.) & 1155 \\
past & at\=ita & (adj.) & \\
present & paccupanna & (adj.) & \\
future & an\=agata & (adj.) & \\
early morning & pacc\=usa & m. & 68 \\
morning & pabh\=ata, pubba\d nha & m. & 68 \\
forenoon & pubba\d nha & m. & \\
midday & majjha\d nha & m. & \\
afternoon & apara\d nha & m. & \\
evening & s\=aya\~nha, dinaccaya & m. & 68 \\
evening & sa\~njh\=a & f. & 68 \\
night & ratti & f. & 69 \\
midnight & nis\=itha & m. & 70 \\
season, weather & utu & m. f. & 79 \\
spring & vasanta & m. & 79 \\
rainy season & vass\=ana & m. & 80--1 \\
summer & gimh\=ana & m. & 80 \\
winter & hemanta & m. & 79 \\
day & divasa, v\=asara, \texthl{aha}\footnote{\pali{Aha} has irregular declension as \pali{mana} group, see page \pageref{decl:mana}.} & m. & 67 \\
day & dina & nt. & 67 \\
week & satt\=aha, sattaratta & nt. & \\
Sunday & raviv\=ara & m. & \\
Monday & candav\=ara & m. & \\
Tuesday & kujav\=ara & m. & \\
Wednesday & budhav\=ara & m. & \\
Thursday & guruv\=ara & m. & \\
Friday & sukkav\=ara & m. & \\
Saturday & saniv\=ara & m. & \\
month & m\=asa & m. & 1110 \\
January & phussa & m. & 76 \\
February & m\=agha & m. & 76 \\
March & phaggu\d na & m. & 76 \\
April & cittam\=asa & m. & 75 \\
May & ves\=akha & m. & 75 \\
June & je\d t\d tham\=asa & m. & 75 \\
July & \=as\=a\d lha & m. & 75 \\
August & s\=ava\d na & m. & 75 \\
September & po\d t\d thap\=ada & m. & 75 \\
October & assayuja & m. & 75 \\
November & kattika & m. & 75 \\
December & m\=agasira & m. & 76 \\
year & vassa, sa\d mvacchara & m. nt. & 81 \\
\end{vocabNtable}

\begin{vocabNtable}{Other common nouns}\label{vocabgrp14}%
army & cam\=u, sen\=a & f. & 381 \\
art, craft & sippa & nt. & 528 \\
beginning & \mbox{pubba, agga, pa\d thama} & nt. & 715 \\
beginning & \=adi & m. & 715 \\
benefit & hita, attha & nt. & \\
cause, reason & hetu & m. & 91 \\
cause, reason & k\=ara\d na & nt. & 91 \\
command & \=a\d n\=a & f. & 354 \\
death & mara\d na & nt. & 404 \\
disadvantage & \=ad\=inava, dosa & m. & 766 \\
discipline, study & sikkh\=a & f. & \\
disease, illness & roga & m. & 323 \\
doubt & \mbox{ka\.nkh\=a, sa\.nkh\=a, vimati} & f. & 170 \\
effort & uss\=aha, v\=ay\=ama & m. & 156 \\
end, goal & anta & m. & 714--5 \\
end, terminal & os\=ana & nt. & 771 \\
example & ud\=ahara\d na & nt. & 115 \\
figure, image & r\=upa & nt. & 825 \\
fortune, wealth & dhana & nt. & \\
going, journey & gamana & nt. & 395 \\
half & a\d d\d dha & nt. & 53 \\
happiness & sukha & nt. & 87--8 \\
health & \=arogya & nt. & 331 \\
hole & \mbox{chidda, vivara, susira} & nt. & 649 \\
killing, murder & m\=ara\d na, hanna & nt. & 403 \\
knowledge & vijj\=a, pa\~n\~n\=a & f. & 152--4\\
language & bh\=as\=a & f. & 105 \\
love & sineha & m. & 173 \\
manner, appearance & \=ak\=ara & m. & 982 \\
medicine & bhesajja & nt. & 330 \\
meeting & sannip\=ata & m. & \\
member & p\=arisajja & m. & \\
message, letter & sandesa & m. & \\
message, letter & s\=asana & nt. & \\
mind & \texthl{mana}\footnote{\pali{Mana} has irregular declension, see page \pageref{decl:mana}.} & m. & 152 \\
mind & citta, vi\~n\~n\=a\d na & nt. & 152 \\
money & m\=ula & nt. & \\
news, event & pavatti & f. & 113 \\
number & sa\.nkhy\=a, ga\d nan\=a & f. & \\
object, thing, land & vatthu & nt. & 970 \\
part, limb & a\.nga & nt. & 278 \\
path & magga & m. & 190--1 \\
pit & k\=asu & f. & 650 \\
portion & bh\=aga & m. & 485 \\
price, value & aggha & m. & 1048 \\
property, belonging & \=ayatta, santaka & m. & 728 \\
question & pa\~nha & m. & 115 \\
sound & sadda, sara & m. & 128 \\
speech & kath\=a, v\=ac\=a & f. & 105 \\
stoppage, return & nivattana & nt. & \\
substance & dabba & nt. & 913 \\
suffering, pain & dukkha & nt. & 89 \\
summary & sa\.nkhepa, sa\.ngaha & m. & 116 \\
teaching & ov\=ada & m. & 354 \\
thought & vitakka, sa\.nkappa & m. & 155 \\
truth & sacca & nt. & 127 \\
victory & jaya & m. & 402 \\
wage, payment, fee & vetana & nt. & 531 \\
walking, conduct & cara\d na & nt. & \\
war & sa\.ng\=ama & m. & 399 \\
ware, article, good & bha\d n\d da, pa\d niya & nt. & 471 \\
wealth & dhana & nt. & 485 \\
weapon & \=avudha & nt. & 385 \\
work, action & kamma\d m & nt. & 758 \\
\end{vocabNtable}

\section{Adjectives}\label{vocab:adj}

I list adjectives into two tables. The first contains words and their opposites. I think it will be helpful, because when we think of an adjective, we often think of its opposite as well. For the limitation of space, I list only succinct definitions. For nuances of meaning, please consult a dictionary. The second table contains adjectives that have no opposite terms, e.g.\ colors, shapes, and other good-to-knows.

\bigskip
\begin{vocabTCtable}{Common adjectives}
abundant & bahula, bhiyya & 703 \\
\smallindent$\neq$ scarce & virala, dullabha & \\
beautiful, good & sundara & 693--4 \\
\smallindent$\neq$ bad, improper, ugly & asundara & \\
beautiful & sur\=upa & \\
\smallindent$\neq$ ugly & vir\=upa & \\
beautiful, shining & sobhana & 693--4 \\
\smallindent$\neq$ ugly, improper & asobhana & \\
big, great & mahanta & 700--1 \\
\smallindent$\neq$ small, little & khuddaka, paritta & 704--5 \\
brave, bold & nibbhaya, v\=ira & \\
\smallindent$\neq$ fearful, timid & bh\=iruka & 731 \\
bright, shining & bh\=asura, pabhassara & \\
\smallindent$\neq$ dim & nippabha & \\
broad, wide & puthula, vitthata & \\
\smallindent$\neq$ narrow & aputhula & \\
clever, skillful & kusala, pa\d tu, dakkha & 720--1 \\
\smallindent$\neq$ unskillful & b\=ala, akusala & 721 \\
clean & nimmala & \\
\smallindent$\neq$ dirty & kali\d t\d tha, malina & \\
clear, unclouded & accha & 670 \\
\smallindent$\neq$ unclear, clouded &  anaccha & 669 \\
concealed & gutta, pa\d ticchanna & \\
\smallindent$\neq$ exposed, opened & viva\d ta, an\=ava\d ta & \\
deep & gambh\=ira & 669 \\
\smallindent$\neq$ shallow & utt\=ana & 669 \\
desirable & i\d t\d tha & 697 \\
\smallindent$\neq$ undesirable & ani\d t\d tha & \\
dry & sukkha & \\
\smallindent$\neq$ wet & alla, tinta & \\
easy to do & sukara & \\
\smallindent$\neq$ hard to do & dukkara & \\
eternal & sassatika & \\
\smallindent$\neq$ momentary & kha\d nika & \\
even & sama & \\
\smallindent$\neq$ uneven, unequal & visama, asama & \\
exceeding, more & adhika & 712 \\
\smallindent$\neq$ less, deficient & \=una & \\
external & b\=ahira & 703 \\
\smallindent$\neq$ internal & abbhantara & \\
extreme & accanta, parama & \\
\smallindent$\neq$ moderate & majjhima, matta\~n\~n\=u & \\ 
famous, renowned & vissuta, sup\=aka\d ta & 724 \\
\smallindent$\neq$ unknown & ap\=aka\d ta, apa\~n\~n\=ata & \\
fast, quick & s\=igha & 40 \\
\smallindent$\neq$ slow & dandha, manda & \\
fat & th\=ula & 701 \\
\smallindent$\neq$ thin, haggard & kisa & \\
fine, sublime & sukhuma & \\
\smallindent$\neq$ coarse & l\=ukha, o\d l\=arika & \\
fortunate, lucky & \mbox{dha\~n\~na, subhaga, bhadda} & \\
\mbox{\smallindent$\neq$ unfortunate, unlucky} & dukkhita, bh\=agyah\=ina & \\
fragrant & sugandha & 146 \\
\smallindent$\neq$ bad smelling & duggandha & 146 \\
full & pu\d n\d na & 749 \\
\smallindent$\neq$ empty & tuccha & 698 \\
gentle & mudu, komala & \\
\smallindent$\neq$ fierce & ghora, ca\d n\d da, d\=aru\d na & \\
glad, cheerful & sumana & 723 \\
\smallindent$\neq$ sad & dummana & 723 \\
good & bhaddaka, sundara & \\
\smallindent$\neq$ bad & du\d t\d tha & \\
happy & sukhita & \\
\smallindent$\neq$ unhappy & dukkhita & \\
hard & thaddha & \\ 
\smallindent$\neq$ soft & mudu & 716 \\
heavy & garuka, bh\=ariya & \\
\smallindent$\neq$ light & lahuka, sallahuka & 710 \\
high & ucca & \\
\smallindent$\neq$ low & n\=ica & \\
hot & u\d nha & \\
\smallindent$\neq$ cold & s\=ita, s\=itala & 56 \\
honest, straight & uju, ava\.nka & 708 \\ 
\smallindent$\neq$ crooked & ku\d tila, va\.nka & 709 \\
\smallindent$\neq$ bent down, stooped & onata & \\
hostile & viruddha & \\
\smallindent$\neq$ friendly, agreeable & \mbox{aviruddha, suhada, anuk\=ula} & \\
kind & k\=aru\d nika & \\
\smallindent$\neq$ cruel & niddaya & \\
lazy & alasa, tandita & \\
\smallindent$\neq$ diligent & analasa, atandita & \\
long & d\=igha, \=ayata & 707 \\
\smallindent$\neq$ short & rassa & \\
many & bahuka & \\
\smallindent$\neq$ little & thoka & \\
near & \=asanna, sam\=ipa & 705--6 \\
\smallindent$\neq$ far & an\=asanna, d\=ura & 706 \\
new, fresh & nava, abhinava & 713 \\
\smallindent$\neq$ old & pur\=a\d na, ji\d n\d na & 713 \\
noble & pavara, uttama & 694--6 \\
\smallindent$\neq$ ignoble, mean & adhama, h\=ina & 699--700 \\
old & mahallaka & 254 \\
\smallindent$\neq$ young & taru\d na, b\=ala & 252 \\
permanent & nicca & 41, 709 \\ 
\smallindent$\neq$ impermanent & anicca & \\
precious & mahaggha & \\
\smallindent$\neq$ worthless & appaggha & \\
pure & suddha, parisuddha & \\
\smallindent$\neq$ impure & asuddha & \\
rich & \texthl{dhanavantu}\footnote{Terms with color highlighted have irregular declension like \pali{gu\d navantu}, see Appendix \ref{decl:irrn}, page \pageref{decl:gunavm}.} & \\
\smallindent$\neq$ poor & da\d lidda & \\
righteous & pu\~n\~na & 85 \\
\smallindent$\neq$ evil & p\=apa & 84 \\
rough & kakkasa & \\
\smallindent$\neq$ smooth & siniddha, ma\d t\d tha & \\
sharp & tikhi\d na & \\
\smallindent$\neq$ blunt & atikhi\d na & \\
stable, fixed & acala, th\=avara & \\
\smallindent$\neq$ moving, unsteady & cala, ja\.ngama & 712--3 \\
\smallindent$\neq$ loose, lax & sithila & \\
strong & pabala & \\
\smallindent$\neq$ weak & dubbala & \\
thick & ghana & \\
\smallindent$\neq$ thin & tanu & \\
virtuous & sus\=ila, \texthl{gu\d navantu} & \\
\smallindent$\neq$ vicious & duss\=ila & \\
wise & \texthl{pa\~n\~navantu} & \\
\smallindent$\neq$ stupid & dandha & \\
\end{vocabTCtable}

\begin{vocabTCtable}{Miscellaneous adjectives}
azure & mandan\=ila & \\
black & k\=a\d la & 96 \\
blue & n\=ila & 96 \\
brown & pi\.ngala & 98 \\
dark brown & ka\d nhap\=ita & \\
green & harita & 97 \\
grey & dh\=usara & 96 \\
light red & aru\d na & 97 \\
multicolored & kamm\=asa & 99 \\
orange & p\=itaratta & \\
pink & p\=a\d tala & 97 \\
red & ratta, lohita & 95 \\
tawny & kapila & 98 \\
yellow & p\=ita & 97 \\
yellowish white & pa\d n\d duva\d n\d na & 96 \\
white & sukka, od\=ata, seta & 95 \\
circular & cakk\=ak\=ara & \\
oval, egg-shaped & a\d n\d d\=ak\=ara & \\
semi-circular & addhacakk\=ak\=ara & \\
square & caturassa & \\
triangular & tiko\d naka & \\
octagonal & a\d t\d tha\d msa, a\d t\d tha\d msika & \\
round & va\d t\d tula, va\d t\d ta & 707 \\
all, whole, entire & sabba, sakala & 702 \\
each, one by one & ekeka & \\
excellent & pa\d nita, atisundara, visi\d t\d tha, uttama & 694--6 \\
few, some & ekacca, katipaya & \\
half & a\d d\d dha & 53 \\
foreign & videsika, vij\=atika & \\
one's own & saka, sak\=iya & \\
remaining, left & sesa & \\
equal & sama, sam\=ana & 529--30 \\
pleasurable & rama\d n\=iya & \\
suitable, proper & patir\=upa & 715 \\
useless, vain & mogha & 715 \\
various & aneka, vividha, pahuvidha, n\=an\=ar\=upa & 717 \\
enough & paho\d naka & \\
\end{vocabTCtable}

\section{Verbs}\label{vocab:verb}

Words that work as verb are rich in P\=ali. They have several forms. Some are easy to compose, some are not. I list here only four forms of verbs:
(1) dictionary form (present, 3rd-person, singular, active-voiced), (2) aorist (simple past) of dict., (3) past participle, and (4) absolutive. Not every word has all these forms. The absents are those never used in the scriptures. It might seem sensible to build them from the rules we have. So, I have filled a few missing words myself. Those words are safe to add for their regularity. But some are not, so I left them missing. We have around 260 verbs here. Other forms of verbs are, in a way, easy to handle. You can learn them in other parts of the book.

\bigskip
\begin{vocabVtable}{Common verbs}
abuse & akkosati & akkocchi & akku\d t\d tha & akkositv\=a \\
address & \=alapati & \=alapi & \=alapita & \=alapitv\=a \\
ail & rujati & ruji & & rujitv\=a \\
anoint & a\~njati & a\~nji & a\~njita & a\~njitv\=a \\
answer & vissajjeti & vissajjesi & vissa\d t\d tha & \mbox{vissajjetv\=a} \\
appear & pa\~n\~n\=ayati & pa\~n\~n\=ayi & pa\~n\~n\=ata & \mbox{pa\~n\~n\=ayitv\=a} \\
arise & uppajjati & uppajji & uppanna & uppajjitv\=a \\
arise & nibbattati & nibbatti & nibbatta & nibbattitv\=a \\
ascend & \=aruhati & \=aruhi & \=ar\=u\d lha & \=aruyha \\
assemble & sannipatati & sannipati & \mbox{sannipatita} & \mbox{sannipatitv\=a} \\
associate & sevati & sevi & sevita & sevitv\=a \\
associate & bhajati & bhaji & bhajita & bhajitv\=a \\
attain & \mbox{adhigacchati} & \mbox{adhigacchi} & adhigata & adhigantv\=a \\
attempt & ussahati & ussahi & & ussahitv\=a \\
awake & pabujjhati & pabujjhi & pabuddha & \mbox{pubujjhitv\=a} \\
awaken & pabodheti & \mbox{pabodhesi} & pabodhita & \mbox{pabodhetv\=a} \\
balance & tuleti & tulesi & tulita & tuletv\=a \\
bathe & nah\=ayati & nah\=ayi & nah\=ata & nah\=ayitv\=a \\
be, become & bhavati & bhavi & bh\=uta & bhavitv\=a \\
be, become & hoti & ahosi & & hutv\=a \\
be able & sakkoti & asakkhi & & sakku\d nitv\=a \\
be angry & kuppati & kuppi & kuppita & kuppitv\=a \\
be established & pati\d t\d th\=ati & pati\d t\d th\=asi & pati\d t\d thita & pati\d t\d th\=aya \\
be fond of & piy\=ayati & piy\=ayi & piy\=ayita & piy\=ayitv\=a \\
be full & \=ap\=urati & \=ap\=uri & \=apu\d n\d na & \=ap\=uriya \\
be manifest & \mbox{p\=atubhavati} & \mbox{p\=atubhavi}, \mbox{p\=aturahosi} & p\=atubh\=uta & \mbox{p\=atubhavitv\=a} \\
be purified & visujjhati & visujjhi & visuddha & visujjhitv\=a \\
be satisfied & tussati & tussi & tu\d t\d tha & tussitv\=a \\
bear & dh\=areti & dh\=aresi & dh\=arita & dh\=aretv\=a \\
beat & paharati & pahari & paha\d ta & paharitv\=a \\
beg & y\=acati & y\=aci & y\=acita & y\=acitv\=a \\
begin & \=arabhati & \=arabhi & \=araddha & \=arabbha \\
believe & saddahati & saddahi & saddahita & saddahitv\=a \\
believe & vissasati & vissasasi & vissattha & vissasitv\=a \\
bind & bandhati & bandhi & bandha & bandhiya \\
bite & \d dasati & \d dasi & \d da\d t\d tha & \d dasitv\=a \\
blow (wind) & v\=ayati & v\=ayi & v\=ayita & v\=ayitv\=a \\
blow (mouth) & dhamati & dhami & dhanta & dhamitv\=a \\
break & bhindati & bhindi & bhindita & bhinditv\=a \\
break & bha\~njati & bha\~nji & bha\~njita & bha\~njitv\=a \\
bring & \=aharati & \=ahari & \=aha\d ta & \=aharitv\=a \\
bring up & poseti & posesi & posita & posetv\=a \\
bury & nikha\d nati & nikha\d ni & nikh\=ata & nikha\d nitv\=a \\
burn (v.i.) & \d dayhati & \d dayhi & da\d d\d dha & \d dayhitv\=a \\
burn (v.t.) & dahati & dahi & da\d d\d dha & dahitv\=a \\
buy & ki\d n\=ati & ki\d ni & k\=ita & ki\d nitv\=a \\
carry & harati & ah\=asi & ha\d ta & haritv\=a \\
carry & vahati & vahi & vahita & vahitv\=a \\
cease & nivattati & nivatti & nivatta & nivattiya \\
cease & nirujjhati & nirujjhi & niruddha & nirujjhitv\=a \\
collect & oci\d n\=ati & ocini & ocita & ocinitv\=a \\
collect & upacin\=ati & upacini & upacita & upacinitv\=a \\
combine & sa\d myojeti & \mbox{sa\d myojesi} & sa\d myutta & \mbox{sa\d myojetv\=a} \\
come & \=agacchati & \=agacchi & \=agata & \=agantv\=a, \=agamma \\
compose & racayati & racayi & racita & racitv\=a \\
conceal & \mbox{pa\d ticch\=adeti} & \mbox{pa\d ticch\=adesi} & \mbox{pa\d ticchanna} & \mbox{pa\d ticch\=adiya} \\
concentrate & \mbox{sam\=adahati} & sam\=adahi & sam\=ahita & \mbox{sam\=adahitv\=a} \\
cover & ch\=adeti & ch\=adesi & channa, ch\=adita & ch\=adetv\=a, ch\=adiya \\
cook & pacati & paci & pacita, pakka & pacitv\=a \\
create & m\=apeti & m\=apesi & m\=apita & m\=apetv\=a \\
create & nimmi\d n\=ati & nimmi\d ni & nimmita & \mbox{nimmi\d nitv\=a} \\
cry & rodati & rodi & rodita & roditv\=a \\
cross & tarati & tari & ti\d n\d na & taritv\=a \\
cure & tikicchati & tikicchi & tikicchita & tikicchiya \\
cut & chindati & chindi & chinna & chinditv\=a \\
dance & naccati & nacci & & naccitv\=a \\
decorate & ma\d n\d deti & ma\d n\d desi & ma\d n\d dita & ma\d n\d detv\=a \\
decorate & ala\.nkaroti & ala\.nkari & ala\.nkata & \mbox{ala\.nkaritv\=a} \\
defeat & par\=ajeti & par\=ajesi & par\=ajita & par\=ajetv\=a \\
delight in & ramati & rami & rata & ramitv\=a \\
deprive of & voropeti & voropesi & voropita & voropetv\=a \\
descend & oruhati & oruhi & or\=u\d lha & oruhitv\=a, oruyha \\
destroy & n\=aseti & n\=asesi & n\=asita & n\=asetv\=a \\
destroy & vin\=aseti & vin\=asesi & vin\=asita & vin\=asetv\=a \\
die & marati & mari & mata & maritv\=a \\
digest & j\=ireti & j\=iresi & ji\d n\d na & j\=iretv\=a \\
dig & kha\d nati & kha\d ni & khata & kha\d nitv\=a \\
disappear & antaradh\=a\-yati & antaradh\=a\-yi & antarahita & antaradh\=a\-yitv\=a \\
dispel & panudati & panudi & panudita & panuditv\=a \\
dispute & vivadati & vivadi & vivadita & vivaditv\=a \\
divide & bh\=ajeti & bh\=ajesi & bh\=ajita & bh\=ajetv\=a \\
divide & vibhajati & vibhaji & vibhatta & vibhajitv\=a \\
do & karoti & kari, ak\=asi & kata & katv\=a, karitv\=a \\
drag & \=aka\d d\d dhati, \=akassati & \=akassi & \=aka\d d\d dhita & \=aka\d d\d dhitv\=a \\
dress (v.t.) & acch\=adeti & \mbox{acch\=adesi} & acch\=adita & acch\=aditv\=a \\
dress (v.i.) & niv\=aseti & niv\=asesi & nivattha & niv\=asetv\=a \\
drive & p\=ajeti & p\=ajesi & p\=ajita & p\=ajetv\=a \\
drive & s\=areti & s\=aresi & s\=arita & s\=aretv\=a \\
dry (v.t.) & sukkh\=apeti & sukkh\=apesi & sukkh\=apita & sukkh\=apetv\=a \\
dry (v.i.) & sussati & sussi & sukkha & sussitv\=a \\
dwell & vasati & vasi & vuttha & vasitv\=a \\
dwell & viharati & vihari & & viharitv\=a \\
dwindle & parih\=ayati & parih\=ayi & parih\=ina & \mbox{parih\=ayitv\=a} \\
eat & bhu\~njati & bhu\~nji & bhutta & bu\~njitv\=a \\
eat & asn\=ati & & asita & \\
eat & adati & adi & & aditv\=a \\
emerge\footnote{from water} & uttarati & udat\=ari & utti\d n\d na & uttaritv\=a \\
endure & sahati & sahi & sahita & sahitv\=a \\
endure & adhiv\=aseti & \mbox{adhiv\=asesi} & \mbox{adhiv\=asita} & \mbox{adhiv\=asetv\=a} \\
enter & pavisati & pavisi & pavi\d t\d tha & pavisitv\=a \\
establish & pati\d t\d th\=apeti & pati\d t\d th\=apesi & pati\d t\d th\=apita & pati\d t\d th\=apiya \\
faint & mucchati & mucchi & mucchita & mucchitv\=a \\
fall & patati & pati & patita & patitv\=a \\
fear & bh\=ayati & bh\=ayi & bh\=ita & bh\=ayitv\=a \\
feel & vedeti & vedesi & vedita & vedetv\=a \\
fight & yujjhati & yujjhi & yujjhita & yujjhitv\=a \\
fill (v.t.) & p\=ureti & p\=uresi & p\=urita & p\=uretv\=a \\
fill (v.i.) & p\=urati & p\=uri & pu\d n\d na & p\=uritv\=a \\
find & upalabhati & upalabhi & \mbox{upaladdha} & upalabbha \\
flee & pal\=ayati & pal\=ayi & pal\=ata & pal\=ayitv\=a \\
fly & uppatati & uppati & uppatita & uppatitv\=a \\
float (v.t.) & plavati & plavi & plavita & plavitv\=a \\
float (v.i.) & vuyhati & vuyhi & v\=u\d lha & vuyhitv\=a \\
flow & paggharati & pagghari & \mbox{paggharita} & \mbox{paggharitv\=a} \\
flow & sandati & sandi & sandita & sanditv\=a \\
fold & sa\d mharati & sa\d mhari & sa\d mha\d ta & \mbox{sa\d mharitv\=a} \\
follow & \mbox{anugacchati} & \mbox{anugacchi} & anugata & anugantv\=a \\
forget & pamussati & pamussi & pamu\d t\d tha & \mbox{pamussitv\=a} \\
forsake & cajati & caji & catta & cajitv\=a \\
forsake & pajahati & pajahi & pajahita & pah\=aya \\
forsake & jahati & jahi & jahita & jahitv\=a \\
frighten & tajjeti & tajjesi & tajjita & tajjetv\=a \\
get & labhati & labhi & laddha & labhitv\=a, laddh\=a \\
give & dad\=ati, deti & ad\=asi & dinna & daditv\=a, datv\=a \\
go & gacchati & gacchi, agam\=asi & gata & gantv\=a \\
go & y\=ati & & y\=ata & y\=atv\=a \\
go beyond & atikkamati & atikkami & atikkanta & \\
go out & nikkhamati & nikkhami & nikkhanta & nikkhamma \\
grow (v.t.) & va\d d\d dheti & va\d d\d dhesi & va\d d\d dhita & va\d d\d dhetv\=a \\
grow (v.i.) & va\d d\d dhati & & vu\d d\d dha & va\d d\d dhitv\=a \\
grow up & r\=uhati & r\=uhi & r\=u\d lha & r\=uhitv\=a \\
guard (conceal) & gopeti & gopesi & gutta & gopetv\=a \\
hang (v.i.) & olambati & olambi & olambita & olambitv\=a \\
hear & su\d n\=ati, su\d noti & su\d ni & suta & su\d nitv\=a, sutv\=a \\
honor & p\=ujeti & p\=ujesi & p\=ujita & p\=ujetv\=a \\
honor & m\=aneti & m\=anesi & m\=anita & m\=anetv\=a \\
illuminate & obh\=aseti & obh\=asesi & obh\=asita & obh\=asetv\=a \\
illuminate & joteti & jotesi & jotita & jotetv\=a \\
interrogate & pucchati & pucchi & pucchita, pu\d t\d tha & pucchitv\=a \\
interrupt & \mbox{upacchindati} & \mbox{upacchindi} & \mbox{upacchinna} & upacchijja \\
judge & t\=ireti & t\=iresi & t\=irita & t\=iretv\=a \\
judge & viniccheti & \mbox{vinicchesi} & vinicchita & \mbox{vinicchetv\=a} \\
jump & \mbox{pakkhandati} & pakkhandi & pakkhanta & \mbox{pakkhanditv\=a} \\
kill & gh\=ateti & gh\=atesi & gh\=atita & gh\=atetv\=a \\
kill & m\=areti & m\=aresi & m\=arita & m\=aretv\=a \\
kill & hanati & hani & hata & hantv\=a \\
kindle & j\=aleti & j\=alesi & j\=alita & j\=aletv\=a \\
knit & gantheti & ganthesi & ganthita & ganthetv\=a \\
know & j\=an\=ati & j\=ani & \~n\=ata & \~n\=atv\=a \\
know & bujjhati & bujjhi & buddha & bujjhitv\=a \\
lament & vilapati & vilapi & & vilapitv\=a \\
laugh & hasati & hasi & hasita & hasitv\=a \\
lay down & nikkhipati & nikkhipi & nikkhitta & \mbox{nikkhipitv\=a} \\
lead & neti & nesi & n\=ita & netv\=a \\
learn & ugga\d nh\=ati & ugga\d nhi & uggahita & \mbox{ugga\d nhitv\=a} \\
learn & sikkhati & sikkhi & sikkhita & sikkhitv\=a \\
lift up & ucc\=areti & ucc\=aresi & ucc\=arita & ucc\=aretv\=a \\
lift up & ukkhipati & ukkhipi & ukkhitta & ukkhipitv\=a \\
look & oloketi & olokesi & olokita & oloketv\=a \\
lose & virajjhati & virajjhi & viraddha & virajjhitv\=a \\
measure & mi\d n\=ati & mi\d ni & mita & mi\d nitv\=a \\
move & calati & cali & calita & calitv\=a \\
observe\footnote{to fulfil a commitment, to comply with, e.g.\ to observe the law} & pa\d tipajjati & pa\d tipajji & pa\d tipanna & \mbox{pa\d tipajjitv\=a} \\
open & vivarati & vivari & viva\d ta & vivaritv\=a \\
oppress & p\=i\d leti & p\=i\d lesi & p\=i\d lita & p\=i\d letv\=a \\
perish & nassati & nassi & na\d t\d tha & nassitv\=a \\
perish & vinassati & vinassi & vina\d t\d tha & vinassitv\=a \\
place & \d thapeti & \d thapesi & \d thapita & \d thapetv\=a \\
play & k\=i\d lati & k\=i\d li & k\=i\d lita & k\=i\d litv\=a \\
plough & kasati & kasi & ka\d t\d tha & kasitv\=a \\
pound & ko\d t\d teti & ko\d t\d tesi & ko\d t\d tita & ko\d t\d tetv\=a \\
praise & \mbox{abhitthavati} & abhitthavi & \mbox{abhitthutta} & \mbox{abhitthaviya} \\
preach & deseti & desesi & desita & desetv\=a \\
prepare & sajjeti & sajjesi & sajjita & sajjetv\=a \\
prevent & niv\=areti & niv\=aresi & niv\=arita & niv\=aretv\=a \\
promise & pa\d tij\=anati & pa\d tij\=ani & pa\d ti\~n\~n\=ata & \mbox{pa\d tij\=anitv\=a} \\
proceed & pavattati & & pavatta, pavattita & pavattetv\=a \\
produce & \=ap\=adeti & \=ap\=adesi & \=ap\=adita & \=ap\=adetv\=a \\
produce & upp\=adeti & upp\=adesi & upp\=adita & upp\=adetv\=a \\
produce & nibbatteti & \mbox{nibbattesi} & nibbattita & \mbox{nibbattetv\=a} \\
protect & rakkhati & rakkhi & rakkhita & rakkhitv\=a \\
protect & p\=aleti & p\=alesi & p\=alita & p\=aletv\=a \\
purify & visodheti & visodhesi & visodhita & visodhetv\=a \\
purge & vireceti & virecesi & virecita & virecetv\=a \\
rain & vassati & vassi & va\d t\d tha & vassitv\=a \\
raise & u\d t\d th\=apeti & u\d t\d th\=apesi & u\d t\d th\=apita & u\d t\d th\=apetv\=a \\
read & pa\d thati & pa\d thi & pa\d thita & pa\d thitv\=a \\
recite & sajjh\=ayati & sajjh\=ayi & sajjh\=ayita & \mbox{sajjh\=ayitv\=a} \\
remember & anussarati & anussari & \mbox{anussarita} & \mbox{anussaritv\=a} \\
resolve & adhi\d t\d th\=ati & \mbox{adhi\d t\d th\=asi} & adhi\d t\d thita & adhi\d t\d th\=aya \\
return (v.i.) & \mbox{pa\d tinivattati} & \mbox{pa\d tinivatti} & \mbox{pa\d tinivatta} & pa\d tinivatti\-tv\=a \\
return (v.i.) & \mbox{pacc\=agacchati} & \mbox{pacc\=agacchi} & pacc\=agata & \mbox{pacc\=agantv\=a} \\
rise & u\d t\d th\=ati & u\d t\d th\=asi & u\d t\d thita & u\d t\d th\=aya \\
roar & nadati & nadi & nadita & naditv\=a \\
run & dh\=avati & dh\=avi & dh\=avita & dh\=avitv\=a \\
safeguard & anup\=aleti & anup\=alesi & anup\=alita & anup\=aletv\=a \\
satisfy & ruccati & rucci & ruccita & ruccitv\=a \\
say & katheti & kathesi & kakhita & kathetv\=a \\
say & bh\=asati & abh\=asi & bh\=asita & bh\=asitv\=a \\
say & vadati & vadi & vutta & vaditv\=a, vatv\=a \\
scatter & vikirati & vikiri & viki\d n\d na & vikiritv\=a \\
see & passati & passi & di\d t\d tha & passitv\=a, disv\=a \\
seek & gavesati & gavesi & gavesita & gavesitv\=a \\
seek & pariyesati & pariyesi & pariyesita & \mbox{pariyesitv\=a} \\
sell & vikki\d n\=ati & vikki\d ni & vikkita & vikki\d nitv\=a \\
send & peseti & pesesi & pesita & pesetv\=a \\
sew & sibbati & sibbi & sibbita & sibbitv\=a \\
shake & kampeti & kampesi & kampita & kampetv\=a \\
shake & c\=aleti & c\=alesi & c\=alita & c\=aletv\=a \\
shine & virocati & viroci & virocita & virocitv\=a \\
shine & obh\=asati & obh\=asi & obh\=asita & obh\=asitv\=a \\
shoot & vijjhati & vijjhi & viddha & vijjhitv\=a \\
show & dasseti & dassesi & dassita & dassetv\=a \\
shut & pidahati & pidahi & pihita & pidahitv\=a \\
sing & g\=ayati & g\=ayi & \mbox{g\=ayita, g\=ita} & g\=ayitv\=a \\
sink & nimujjati & nimujji & nimugga & nimujjitv\=a \\
sit & nis\=idati & nis\=idi & nisinna & nis\=iditv\=a, nis\=idiya \\
sleep & \mbox{sayati, seti} & sesi, sayi & say\=apita & sayitv\=a \\
sleep & nidd\=ayati & nidd\=ayi & nidd\=ayita & nidd\=ayitv\=a \\
smell & gh\=ayati & gh\=ayi & gh\=ayita & gh\=ayitv\=a \\
spread (v.t.) & vitth\=areti & \mbox{vitth\=aresi} & vitth\=arita & vitth\=aretv\=a \\
spread (v.i.) & pattharati & patthari & pattha\d ta & \mbox{pattharitv\=a} \\
spring up & ubbhijjati & ubbhijji & ubbhinna & ubbhijja \\
stand & ti\d t\d thati & a\d t\d th\=asi & \d thita & \d thatv\=a \\
steal & coreti & coresi & corita & coretv\=a \\
steal & theneti & thenesi & thenita & thenetv\=a \\
string & \=avu\d n\=ati & \=avu\d ni & \=avuta & \=avu\d nitv\=a \\
stumble & pakkhalati & pakkhali & pakkhalita & \mbox{pakkhalitv\=a} \\
subsist & j\=ivati & j\=ivi & & j\=ivitv\=a \\
succeed & sijjhati & sijjhi & siddha & \\
swallow & gilati & gili & gilita & gilitv\=a \\
sweep & sammajjati & sammajji & \mbox{samma\d t\d tha} & \mbox{sammajjitv\=a} \\
take & \=ad\=ati & & \=adinna & \=ad\=aya \\
take & ga\d nh\=ati & ga\d nhi & gahita & gahetv\=a \\
take down & ot\=areti & ot\=aresi & ot\=arita & ot\=aretv\=a \\
take down & oropeti & oropesi & oropita & oropetv\=a \\
talk & sallapati & sallapi & & sallapitv\=a \\
talk & sambh\=asati & \mbox{samabh\=asi} & \mbox{sambh\=asita} & \mbox{sambh\=asitv\=a} \\
taste & s\=ayati & s\=ayi & s\=ayita & sayitv\=a \\
teach & \mbox{ugga\d nh\=apeti} & \mbox{ugga\d nh\=apesi} & \mbox{ugga\d nh\=apita} & ugga\d nh\=ape\-tv\=a \\
teach & sikkh\=apeti & sikkh\=apesi & sikkh\=apita & sikkh\=apetv\=a \\
tell, inform & \=acikkhati & \=acikkhi & \=acikkhita & \=acikkhitv\=a \\
tell, inform & \=aroceti & \=arocesi & \=arocita & \=arocetv\=a \\
think & cinteti & cintesi & cintita & cintetv\=a \\
throw & khipati & khipi & khitta & khipitv\=a \\
touch & phusati & phusi & phu\d t\d tha & phusitv\=a \\
train & dameti & damesi & damita, danta & dametv\=a \\
translate & parivatteti & \mbox{parivattesi} & parivattita & \mbox{parivattetv\=a} \\
tremble & kampati & kampi & kampita & kampitv\=a \\
turn & va\d t\d teti & va\d t\d tesi & va\d t\d tita & va\d t\d tetv\=a \\
unfold & pas\=areti & pas\=aresi & pas\=arita & pas\=aretv\=a \\
vomit & vamati & vami & vamita & vamitv\=a \\
wait for & \=agameti & \=agamesi & & \=agametv\=a \\
walk & carati & cari & carita & caritv\=a \\
wander & \=ahi\d n\d dati & \=ahi\d n\d di & & \=ahi\d n\d ditv\=a \\
want & icchati & icchi & icchita & icchiya \\
wash & dhovati & dhovi & dhota & dhovitv\=a \\
weave & vin\=ati & vini & vita & vinitv\=a \\
win & jin\=ati & jini & jita & jinitv\=a \\
wish, hope & pattheti & patthesi & patthita & patthetv\=a \\
wrap & ve\d theti & ve\d thesi & ve\d thita & ve\d thetv\=a \\
write & likhati & likhi & likhita & likhitv\=a \\
yoke & yojeti & yojesi & yutta & yojetv\=a \\
\end{vocabVtable}
